---
title: "lagoon_vs_sea"
format: html
---

# Lagoon vs Sea

This document will compare the ITS2 type profiles between Porites lutea in Aldabra lagoon and in the seaward reef:

1. PCoA of Bray-Curtis distances
2. PERMANOVA test and PERMDISP companion test
3. PCoA of C15 profiles (dominant profiles only)
4. Comparing ITS2 type profile distances between lagoon and sea samples

## Load Libraries

```{r}
library(here)
library(ggplot2)
library(ggpubr)
library(dplyr)
library(vegan)
library(ggrepel)
library(cowplot)
```

# 1. PCoA of Bray-Curtis Distances

```{r}
#| fig-width: 7
#| fig-height: 6
#| dpi: 300

# Load PCoA coordinates (Bray-Curtis, no square root transformation)
pcoa_coords <- read.csv(
  here("symportal_data/20260116T164045_hoongbe1/between_sample_distances/C/20260117T155228_braycurtis_samples_PCoA_coords_C_no_sqrt.csv"),
  row.names = 1
)

# Extract sample names and create location variable
pcoa_coords$sample_name <- rownames(pcoa_coords)

# Assign sampling site based on sample prefix
pcoa_coords$site <- ifelse(grepl("^ARM01", pcoa_coords$sample_name), "ARM01", "ARM09")

# Get proportion of variance explained for axes
prop_var <- pcoa_coords[pcoa_coords$sample_name == "proportion_explained", c("PC1", "PC2")]
pcoa_coords <- pcoa_coords[pcoa_coords$sample_name != "proportion_explained", ]

# Convert PC columns to numeric
pcoa_coords$PC1 <- as.numeric(pcoa_coords$PC1)
pcoa_coords$PC2 <- as.numeric(pcoa_coords$PC2)

# Add jitter to lagoon samples (ARM09) on x-axis for better visibility
set.seed(123)  # For reproducibility
jitter_amount <- 0.025  # Adjust this value to control jitter amount
pcoa_coords$PC1_jittered <- pcoa_coords$PC1
lagoon_indices <- pcoa_coords$site == "ARM09"
pcoa_coords$PC1_jittered[lagoon_indices] <- pcoa_coords$PC1[lagoon_indices] + 
  runif(sum(lagoon_indices), -jitter_amount, jitter_amount)

# Create PCoA plot
p_pcoa <- ggplot(pcoa_coords, aes(x = PC1_jittered, y = PC2, fill = site)) +
  geom_point(size = 3, alpha = 0.8, shape = 21, stroke = 1, color = "black") +
  geom_text_repel(
    data = subset(pcoa_coords, sample_name == "ARM09-8"),
    aes(label = "Colony 16"),
    size = 3,
    nudge_x = 0.05,
    min.segment.length = 0,
    nudge_y = 0.005,
    box.padding = 0.6,
    point.padding = 0.4,
    segment.size = 0.3,
    show.legend = FALSE
  ) +
  scale_fill_manual(
    values = c("ARM01" = "#3498db", "ARM09" = "#e74c3c"),
    labels = c("ARM01" = "Seaward reef", "ARM09" = "Lagoon")
  ) + 
  labs(
    x = paste0("PC1 (", round(as.numeric(prop_var$PC1) * 100, 1), "%)"),
    y = paste0("PC2 (", round(as.numeric(prop_var$PC2) * 100, 1), "%)"),
    fill = "Colony origin"
  ) +
  theme_pubr(base_size = 14) +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.position = c(1, 0.05),
    legend.justification = c(1, 0),
    legend.text = element_text(size = 12, vjust = 0.5),
    legend.background = element_rect(color = "black", fill = "white", linewidth = 0.3),
    legend.margin = margin(4, 6, 4, 6),
    legend.key = element_blank(),
    legend.title = element_text(size = 14, face = "bold")
  )

print(p_pcoa)

# Save plot
ggsave(here("figures/pcoa_braycurtis.png"), p_pcoa, width = 7, height = 6, dpi = 300)
ggsave(here("figures/pcoa_braycurtis.svg"), p_pcoa, width = 7, height = 6)
```

# 2. PERMANOVA Test

```{r}
# Load Bray-Curtis distance matrix
dist_matrix <- read.table(
  here("symportal_data/20260116T164045_hoongbe1/between_sample_distances/C/20260117T155228_braycurtis_sample_distances_C_no_sqrt.dist"),
  header = FALSE,
  sep = "\t",
  row.names = 1
)

# Remove the second column (UIDs) and set column names
sample_names <- rownames(dist_matrix)
dist_matrix <- dist_matrix[, -1]  # Remove UID column
colnames(dist_matrix) <- sample_names

# Convert to dist object
dist <- as.dist(dist_matrix)

# Create group vector
group <- ifelse(grepl("^ARM01", sample_names), "ARM01", "ARM09")

# Perform PERMANOVA (adonis2)
permanova_result <- adonis2(dist ~ group, permutations = 999)

# Print results
print(permanova_result)
```

## PERMDISP Test

Test for homogeneity of multivariate dispersions (assumption check for PERMANOVA)

```{r}
# Perform betadisper analysis (using same distance matrix and groups from above)
bd <- betadisper(dist, group)

# Visualise PERMDISP
plot(bd)

# Test for significant differences in dispersion
permdisp_result <- permutest(bd, permutations = 999)

# Print results
print(permdisp_result)

# Show mean distances to centroids for each group
print(bd)
```

# 3. PCoA of C15 Profiles (Dominant Profiles Only)

Visualize phylogenetic distances between C15 profiles in ordination space

```{r}
#| fig-width: 8
#| fig-height: 7
#| dpi: 300

# Load PCoA coordinates for UniFrac distances
pcoa_coords <- read.csv(
  "symportal_data/20260116T164045_hoongbe1/between_profile_distances/C/20260117T155228_unifrac_profiles_PCoA_coords_C_no_sqrt.csv",
  row.names = 1
)

# Create profile name column from row names
pcoa_coords$profile_name <- rownames(pcoa_coords)

# Extract proportion of variance explained
prop_var <- pcoa_coords[pcoa_coords$profile_name == "proportion_explained", c("PC1", "PC2")]
pcoa_coords <- pcoa_coords[pcoa_coords$profile_name != "proportion_explained", ]

# Filter to only C15 profiles (exclude C91 and C91h)
c15_profiles <- !grepl("^C91", pcoa_coords$profile_name)
excluded_profiles <- pcoa_coords$profile_name[!c15_profiles]
pcoa_coords_c15 <- pcoa_coords[c15_profiles, ]

# Convert PC columns to numeric
pcoa_coords_c15$PC1 <- as.numeric(pcoa_coords_c15$PC1)
pcoa_coords_c15$PC2 <- as.numeric(pcoa_coords_c15$PC2)

# Load sample-profile mapping to classify profiles by habitat
sample_profiles <- read.csv(
  "symportal_data/20260116T164045_hoongbe1/its2_type_profiles/its2_type_profiles_absolute.csv",
  row.names = 1,
  check.names = FALSE
)

# Define habitat-based sample groups
seaward_samples <- rownames(sample_profiles)[grepl("^ARM01", rownames(sample_profiles))]
lagoon_samples <- rownames(sample_profiles)[grepl("^ARM09", rownames(sample_profiles))]

# Classify each profile based on where it's found
# A profile is "Seaward reef" if found only in seaward reef samples
# "Lagoon" if found only in lagoon samples
# "Both" if found in both lagoon and seaward reef samples
classify_profile <- function(profile_name) {
  if (!profile_name %in% colnames(sample_profiles)) {
    return("Unknown")
  }
  
  # Get samples where this profile is present (abundance > 0)
  present_samples <- rownames(sample_profiles)[sample_profiles[, profile_name] > 0]
  
  in_seaward <- any(present_samples %in% seaward_samples)
  in_lagoon <- any(present_samples %in% lagoon_samples)
  
  if (in_seaward && in_lagoon) {
    return("Both")
  } else if (in_seaward) {
    return("Seaward reef")
  } else if (in_lagoon) {
    return("Lagoon")
  } else {
    return("Unknown")
  }
}

# Apply classification to all C15 profiles
pcoa_coords_c15$habitat_status <- sapply(pcoa_coords_c15$profile_name, classify_profile)

cat("=== PCoA of C15 Profiles ===\n")
cat("Total profiles (after removing proportion_explained):", nrow(pcoa_coords), "\n")
cat("C15 profiles plotted:", nrow(pcoa_coords_c15), "\n")
cat("Excluded profiles:", paste(excluded_profiles, collapse = ", "), "\n\n")

cat("Profiles by habitat status:\n")
print(table(pcoa_coords_c15$habitat_status))

# Create PCoA plot with habitat status
p_pcoa_profiles <- ggplot(pcoa_coords_c15, aes(x = PC1, y = PC2, fill = habitat_status)) +
  geom_point(size = 3.5, alpha = 0.8, shape = 24, color = "black", stroke = 1.2) +
  geom_text_repel(aes(label = profile_name), size = 4, max.overlaps = 20, nudge_x = 0,
                  box.padding = 0.5, point.padding = 0.5, min.segment.length = 0.3,
                  bg.color = "white", bg.r = 0.1, segment.size = 0.3) +
  scale_fill_manual(
    values = c("Seaward reef" = "#3498db", "Lagoon" = "#e74c3c", "Both" = "#f39c12"),
    labels = c("Lagoon" = "Lagoon only", 
               "Seaward reef" = "Seaward reef only",
               "Both" = "Both sites")
  ) +
  guides(fill = guide_legend(reverse = TRUE)) +
  labs(
    x = paste0("PC1 (", round(as.numeric(prop_var$PC1) * 100, 1), "%)"),
    y = paste0("PC2 (", round(as.numeric(prop_var$PC2) * 100, 1), "%)"),
    fill = "Profile distribution"
  ) +
  theme_pubr(base_size = 14) +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.position = c(1, 0.05),
    legend.justification = c(1, 0),
    legend.text = element_text(size = 14, vjust = 0.5),
    legend.background = element_rect(color = "black", fill = "white", linewidth = 0.3),
    legend.margin = margin(4, 6, 4, 6),
    legend.key = element_blank(),
    legend.title = element_text(size = 14, face = "bold")
  )

print(p_pcoa_profiles)

# Save plot
ggsave("figures/pcoa_profiles_c15_unifrac.png", p_pcoa_profiles, width = 8, height = 7, dpi = 300)
ggsave("figures/pcoa_profiles_c15_unifrac.svg", p_pcoa_profiles, width = 8, height = 7)
```

## Combined PCoA Panels

```{r}
#| fig-width: 15
#| fig-height: 7
#| dpi: 300

p_pcoa_combined <- plot_grid(
  p_pcoa,
  NULL,
  p_pcoa_profiles,
  labels = c("A", "", "B"),
  ncol = 3,
  rel_widths = c(1, 0.12, 1),
  align = "h"
)

print(p_pcoa_combined)

ggsave("figures/pcoa_combined_panels.png", p_pcoa_combined, width = 15, height = 7, dpi = 300)
ggsave("figures/pcoa_combined_panels.svg", p_pcoa_combined, width = 15, height = 7)
```

# 4. Calculating Bray-Curtis between samples

```{r}
#| fig-width: 7
#| fig-height: 6
#| dpi: 300

# Load post-MED sequence relative abundance table
seq_rel_abund <- read.csv(
  here("symportal_data/20260116T164045_hoongbe1/post_med_seqs/seq_relative_abundance.csv"),
  row.names = 1,
  check.names = FALSE
)

# Calculate Bray-Curtis dissimilarity between samples
seq_bray <- vegdist(seq_rel_abund, method = "bray")

# Run PCoA
seq_pcoa <- cmdscale(seq_bray, k = 2, eig = TRUE)
var_explained <- seq_pcoa$eig / sum(seq_pcoa$eig[seq_pcoa$eig > 0])

# Build coordinates table
pcoa_coords_seq <- as.data.frame(seq_pcoa$points)
colnames(pcoa_coords_seq) <- c("PC1", "PC2")
pcoa_coords_seq$sample_name <- rownames(pcoa_coords_seq)

# Assign sampling site based on sample prefix
pcoa_coords_seq$site <- ifelse(grepl("^ARM01", pcoa_coords_seq$sample_name), "ARM01", "ARM09")

# Add jitter to lagoon samples (ARM09) on x-axis for better visibility
set.seed(1)
jitter_amount <- 0.025
pcoa_coords_seq$PC1_jittered <- pcoa_coords_seq$PC1
lagoon_indices <- pcoa_coords_seq$site == "ARM09"
pcoa_coords_seq$PC1_jittered[lagoon_indices] <- pcoa_coords_seq$PC1[lagoon_indices] +
  runif(sum(lagoon_indices), -jitter_amount, jitter_amount)

# Create PCoA plot in same style as p_pcoa
p_pcoa_seq <- ggplot(pcoa_coords_seq, aes(x = PC1_jittered, y = PC2, fill = site)) +
  geom_point(size = 3.5, alpha = 0.8, shape = 21, stroke = 1, color = "black") +
  geom_text_repel(
    data = subset(pcoa_coords_seq, sample_name == "ARM09-8"),
    aes(label = "Colony 16"),
    size = 4,
    nudge_x = 0.05,
    min.segment.length = 0,
    nudge_y = 0.005,
    box.padding = 0.6,
    point.padding = 0.4,
    segment.size = 0.3,
    show.legend = FALSE
  ) +
  scale_fill_manual(
    values = c("ARM01" = "#3498db", "ARM09" = "#e74c3c"),
    labels = c("ARM01" = "Seaward reef", "ARM09" = "Lagoon")
  ) +
  labs(
    x = paste0("PC1 (", round(var_explained[1] * 100, 1), "%)"),
    y = paste0("PC2 (", round(var_explained[2] * 100, 1), "%)"),
    fill = "Colony origin"
  ) +
  theme_pubr(base_size = 14) +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12),
    legend.position = c(1, 0.05),
    legend.justification = c(1, 0),
    legend.text = element_text(size = 14, vjust = 0.5),
    legend.background = element_rect(color = "black", fill = "white", linewidth = 0.3),
    legend.margin = margin(4, 6, 4, 6),
    legend.key = element_blank(),
    legend.title = element_text(size = 14, face = "bold")
  )

print(p_pcoa_seq)

ggsave(here("figures/pcoa_all_seq_braycurtis.png"), p_pcoa_seq, width = 7, height = 6, dpi = 300)
ggsave(here("figures/pcoa_all_seq_braycurtis.svg"), p_pcoa_seq, width = 7, height = 6)
```

## Combined PCoA Panels

```{r}
#| fig-width: 15
#| fig-height: 7
#| dpi: 300

p_pcoa_combined <- plot_grid(
  p_pcoa_seq,
  NULL,
  p_pcoa_profiles,
  labels = c("A", "", "B"),
  ncol = 3,
  rel_widths = c(1, 0.025, 1),
  align = "h"
)

print(p_pcoa_combined)

ggsave("figures/pcoa_all_seq_combined.png", p_pcoa_combined, width = 15, height = 7, dpi = 300)
ggsave("figures/pcoa_all_seq_combined.svg", p_pcoa_combined, width = 15, height = 7)
```

# 5. Rerun PERMANOVA and PERMDISP for all sequences

```{r}
# Use the Bray-Curtis distance matrix calculated from post-MED sequences in section 4
seq_sample_names <- labels(seq_bray)
seq_group <- ifelse(grepl("^ARM01", seq_sample_names), "ARM01", "ARM09")

# Perform PERMANOVA (adonis2)
seq_permanova_result <- adonis2(seq_bray ~ seq_group, permutations = 999)
print(seq_permanova_result)
```

## PERMDISP Test

```{r}
# Perform betadisper analysis
seq_bd <- betadisper(seq_bray, seq_group)

# Visualise PERMDISP
plot(seq_bd)

# Test for significant differences in dispersion
seq_permdisp_result <- permutest(seq_bd, permutations = 999)
print(seq_permdisp_result)

# Show mean distances to centroids for each group
print(seq_bd)
```

# 6. Comparing ITS2 type profiles between lagoon and sea samples

Are seaward reef profiles more different from each other than lagoon profiles? Short answer: yes observationally but no statistically

### Load Profile Data

```{r}
# Load sample-profile mapping (keep original column names with dashes)
sample_profiles <- read.csv(
  here("symportal_data/20260116T164045_hoongbe1/its2_type_profiles/its2_type_profiles_absolute.csv"),
  row.names = 1,
  check.names = FALSE  # Preserve original column names with dashes
)

# Filter to only Clade C profiles (exclude Clade D)
# Identify which profiles are Clade C vs D based on column names
clade_c_profiles <- grep("^C", colnames(sample_profiles), value = TRUE)
sample_profiles <- sample_profiles[, clade_c_profiles]

cat("Sample-profile data dimensions (Clade C only):", dim(sample_profiles), "\n")
cat("Number of Clade C profiles:", ncol(sample_profiles), "\n")
cat("Profile names (first 10):\n")
print(head(colnames(sample_profiles), 10))
cat("\nFirst few samples and profiles:\n")
print(sample_profiles[1:min(6, nrow(sample_profiles)), 1:min(6, ncol(sample_profiles))])

# Create location mapping
sample_locations <- data.frame(
  sample = rownames(sample_profiles),
  location = ifelse(grepl("^ARM01", rownames(sample_profiles)), "Seaward Reef", "Lagoon")
)

cat("\nSamples per location:\n")
print(table(sample_locations$location))
cat("\nNote: Analysis limited to Clade C profiles only\n")
```

### Load Profile Distance Matrix

```{r}
# Load UniFrac distance matrix between profiles
profile_dist_matrix <- read.table(
  here("symportal_data/20260116T164045_hoongbe1/between_profile_distances/C/20260117T155228_unifrac_profile_distances_C_no_sqrt.dist"),
  header = FALSE,
  sep = "\t",
  row.names = 1
)

# Clean up: remove UID column (column 1) and convert to numeric matrix
profile_names <- rownames(profile_dist_matrix)
profile_dist_matrix <- as.matrix(profile_dist_matrix[, -1])
colnames(profile_dist_matrix) <- profile_names
rownames(profile_dist_matrix) <- profile_names

# Convert to distance object for later use
profile_dist <- as.dist(profile_dist_matrix)

cat("Profile distance matrix loaded\n")
cat("Distance matrix dimensions:", nrow(profile_dist_matrix), "x", ncol(profile_dist_matrix), "\n")
cat("Profile names (first 10):", paste(colnames(profile_dist_matrix)[1:min(10, ncol(profile_dist_matrix))], collapse = ", "), "\n")
```

### Identify Profiles Present at Each Location

```{r}
# For each location, find which profiles are present (abundance > 0)
seaward_samples <- sample_locations$sample[sample_locations$location == "Seaward Reef"]
lagoon_samples <- sample_locations$sample[sample_locations$location == "Lagoon"]

# Get profiles present in each location (any sample with abundance > 0)
seaward_profiles <- colnames(sample_profiles)[
  colSums(sample_profiles[seaward_samples, , drop = FALSE] > 0) > 0
]
lagoon_profiles <- colnames(sample_profiles)[
  colSums(sample_profiles[lagoon_samples, , drop = FALSE] > 0) > 0
]

# Check which profiles exist in the distance matrix
profiles_in_dist <- colnames(profile_dist_matrix)

# Filter to only profiles that exist in distance matrix
seaward_profiles <- seaward_profiles[seaward_profiles %in% profiles_in_dist]
lagoon_profiles <- lagoon_profiles[lagoon_profiles %in% profiles_in_dist]

cat("Seaward Reef:\n")
cat("  Number of profiles:", length(seaward_profiles), "\n")
cat("  Profile names:", paste(seaward_profiles, collapse = ", "), "\n\n")

cat("Lagoon:\n")
cat("  Number of profiles:", length(lagoon_profiles), "\n")
cat("  Profile names:", paste(lagoon_profiles, collapse = ", "), "\n\n")

# Shared vs unique profiles
shared_profiles <- intersect(seaward_profiles, lagoon_profiles)
unique_seaward <- setdiff(seaward_profiles, lagoon_profiles)
unique_lagoon <- setdiff(lagoon_profiles, seaward_profiles)

cat("Shared profiles:", length(shared_profiles), "\n")
cat("Unique to Seaward Reef:", length(unique_seaward), "\n")
cat("Unique to Lagoon:", length(unique_lagoon), "\n")
```

### Calculate Mean Pairwise Phylogenetic Distance

```{r}
# Function to calculate mean pairwise distance for a set of profiles
calculate_mean_pairwise_distance <- function(profiles, dist_matrix) {
  # Convert to matrix if needed
  if (inherits(dist_matrix, "dist")) {
    dist_matrix <- as.matrix(dist_matrix)
  }
  
  # Subset distance matrix to only include profiles present
  subset_dist <- dist_matrix[profiles, profiles]
  
  # Extract upper triangle (excluding diagonal)
  pairwise_dists <- subset_dist[upper.tri(subset_dist)]
  
  if (length(pairwise_dists) == 0) {
    return(NA)
  }
  
  return(mean(pairwise_dists))
}

# Calculate for each location (using all profiles present)
seaward_mean_dist <- calculate_mean_pairwise_distance(seaward_profiles, profile_dist_matrix)
lagoon_mean_dist <- calculate_mean_pairwise_distance(lagoon_profiles, profile_dist_matrix)

cat("=== Mean Pairwise Phylogenetic Distance ===\n\n")
cat("Seaward Reef profiles:\n")
cat("  Mean UniFrac distance:", round(seaward_mean_dist, 4), "\n")
cat("  Number of pairwise comparisons:", choose(length(seaward_profiles), 2), "\n\n")

cat("Lagoon profiles:\n")
cat("  Mean UniFrac distance:", round(lagoon_mean_dist, 4), "\n")
cat("  Number of pairwise comparisons:", choose(length(lagoon_profiles), 2), "\n\n")

cat("Difference (Seaward - Lagoon):", round(seaward_mean_dist - lagoon_mean_dist, 4), "\n")
cat("Percent difference:", round((seaward_mean_dist - lagoon_mean_dist) / lagoon_mean_dist * 100, 1), "%\n")
```

### Statistical Test: Permutation Test

```{r}
# Permutation test: randomize profile assignment to locations and compare distances
set.seed(123)
n_perms <- 9999

# Store all profiles
all_profiles <- colnames(sample_profiles)

# Calculate observed difference
obs_diff <- seaward_mean_dist - lagoon_mean_dist

# Run permutation test
perm_diffs <- numeric(n_perms)

for (i in 1:n_perms) {
  # Randomly permute profile assignments to locations
  perm_seaward <- sample(all_profiles, size = length(seaward_profiles), replace = FALSE)
  perm_lagoon <- setdiff(all_profiles, perm_seaward)
  
  # Calculate mean distances for permuted data
  perm_seaward_dist <- calculate_mean_pairwise_distance(perm_seaward, profile_dist_matrix)
  perm_lagoon_dist <- calculate_mean_pairwise_distance(perm_lagoon, profile_dist_matrix)
  
  perm_diffs[i] <- perm_seaward_dist - perm_lagoon_dist
}

# Calculate p-value (two-tailed)
p_value <- mean(abs(perm_diffs) >= abs(obs_diff))

cat("=== Permutation Test Results ===\n\n")
cat("Observed difference (Seaward - Lagoon):", round(obs_diff, 4), "\n")
cat("Number of permutations:", n_perms, "\n")
cat("Two-tailed p-value:", p_value, "\n\n")

if (p_value < 0.05) {
  cat("RESULT: Seaward Reef profiles are SIGNIFICANTLY different in phylogenetic distance\n")
  cat("        compared to Lagoon profiles (p <", p_value, ")\n")
} else {
  cat("RESULT: No significant difference in phylogenetic distance between locations\n")
  cat("        (p =", p_value, ")\n")
}
```

### Visualize Distribution of Distances

```{r}
#| fig-width: 10
#| fig-height: 5
#| dpi: 300

# Create data frame for visualization
perm_results <- data.frame(
  diff = perm_diffs,
  type = "Permuted"
)

obs_results <- data.frame(
  diff = obs_diff,
  type = "Observed"
)

# Plot 1: Histogram of permuted differences
p1 <- ggplot(perm_results, aes(x = diff, fill = type)) +
  geom_histogram(bins = 50, alpha = 0.7, color = "black") +
  geom_vline(xintercept = obs_diff, color = "red", linetype = "dashed", linewidth = 1.5) +
  scale_fill_manual(values = c("Permuted" = "lightblue")) +
  labs(
    title = "Permutation Test: Difference in Mean Pairwise Distances",
    subtitle = "Red line = observed difference",
    x = "Seaward Mean Distance - Lagoon Mean Distance",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Plot 2: Box plot comparing distances
compare_data <- data.frame(
  location = c(rep("Seaward Reef", length(seaward_profiles)), 
               rep("Lagoon", length(lagoon_profiles))),
  profile = c(seaward_profiles, lagoon_profiles),
  mean_dist = c(rep(seaward_mean_dist, length(seaward_profiles)),
                rep(lagoon_mean_dist, length(lagoon_profiles)))
)

p2 <- ggplot(compare_data, aes(x = location, y = mean_dist, fill = location)) +
  geom_col(alpha = 0.7) +
  scale_fill_manual(values = c("Seaward Reef" = "#3498db", "Lagoon" = "#e74c3c")) +
  labs(
    title = "Mean Pairwise Phylogenetic Distance by Location",
    x = "",
    y = "Mean UniFrac Distance"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggarrange(p1, p2, ncol = 2)
```
