---
title: "symportal_output"
format: html
---

# SymPortal Output

This document will be used for the analysis of SymPortal output for the 16 samples of Porites lutea collected from Aldabra.

## Load Libraries

```{r}
library(here)
library(ggplot2)
library(tidyr)
library(ggpubr)
library(jsonlite)
library(cowplot)
library(dplyr)
```

## ITS2 Sequence Absolute Abundance by Sample

```{r}
#| fig-width: 10
#| fig-height: 10
#| dpi: 300

# Load ITS2 sequence absolute abundance data
seq_absolute_abundance <- read.csv(
  here("symportal_data/20260116T164045_hoongbe1/post_med_seqs/seq_absolute_abundance.csv"),
  check.names = FALSE
)

# Load SymPortal sequence colour dictionary
color_dict <- jsonlite::fromJSON(
  here("symportal_data/20260116T164045_hoongbe1/html/color_dict_post_med.json")
)

# Ensure every sequence in the data has a colour; fall back to grey if missing
sequence_names <- setdiff(names(seq_absolute_abundance), "sample_name")
missing_sequences <- setdiff(sequence_names, names(color_dict))
if (length(missing_sequences) > 0) {
  color_dict <- c(color_dict, setNames(rep("#808080", length(missing_sequences)), missing_sequences))
}

# Order sequences by total abundance; reverse to place least abundant at the bottom
seq_totals_abs <- colSums(seq_absolute_abundance[sequence_names], na.rm = TRUE)
seq_order <- rev(names(sort(seq_totals_abs, decreasing = TRUE)))
color_palette <- color_dict[seq_order]

# Build sample order alphabetically
sample_levels <- sort(unique(seq_absolute_abundance$sample_name))

# Create mapping from sample IDs to numeric colony labels for x-axis display
colony_name_map <- setNames(as.character(seq_along(sample_levels)), sample_levels)

# Convert to long format for ggplot2
seq_long <- seq_absolute_abundance %>%
  pivot_longer(cols = -sample_name, 
               names_to = "sequence", 
               values_to = "abundance")
seq_long$sequence <- factor(seq_long$sequence, levels = seq_order)
seq_long$sample_name <- as.character(seq_long$sample_name)
seq_long$sample_name <- factor(seq_long$sample_name, levels = sample_levels)

# Add site grouping for faceting
seq_long <- seq_long %>%
  mutate(site = ifelse(grepl("^ARM01", sample_name), "Seaward Reef Colonies", "Lagoon Colonies"),
         site = factor(site, levels = c("Seaward Reef Colonies", "Lagoon Colonies")))

# Create stacked bar chart
p1 <- ggplot(seq_long, aes(x = sample_name, y = abundance, fill = sequence)) +
  geom_bar(stat = "identity") +
  facet_grid(~site, scales = "free_x", space = "free_x", switch = "x") +
  scale_x_discrete(labels = colony_name_map) +
  scale_fill_manual(values = color_palette, na.translate = FALSE) +
  scale_y_continuous(expand = c(0, 0), labels = scales::label_number(suffix = "k", scale = 1/1000)) +
  theme_pubr(base_size = 14) +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.position = "bottom",
        legend.text = element_text(size = 10),
        legend.title = element_text(face = "bold"),
        legend.key.size = unit(0.4, "cm"),
        legend.box.just = "center",
        legend.margin = margin(0, 0, 0, 0),
        legend.spacing.y = unit(2, "pt"),
        plot.margin = margin(10, 10, 10, 10),
        strip.background = element_blank(),
        strip.text = element_text(size = 16, face = "bold"),
        strip.placement = "outside",
        panel.spacing.x = unit(8, "mm")) +
  guides(fill = guide_legend(ncol = 8, reverse = TRUE, title.position = "top")) +
  labs(x = "", y = "Absolute Abundance", fill = "ITS2 Sequence")

print(p1)

# Save plot
ggsave(here("figures/seq_absolute_abundance.png"), p1, width = 10, height = 10, dpi = 300)
ggsave(here("figures/seq_absolute_abundance.svg"), p1, width = 10, height = 10)

```

## ITS2 Sequence Relative Abundance by Sample

```{r}
#| fig-width: 10
#| fig-height: 10
#| dpi: 300

# Load ITS2 sequence relative abundance data
seq_relative_abundance <- read.csv(
  here("symportal_data/20260116T164045_hoongbe1/post_med_seqs/seq_relative_abundance.csv"),
  check.names = FALSE
)

# Use the same sequence order and colours for consistency
rel_sequence_names <- setdiff(names(seq_relative_abundance), "sample_name")
# Align palette to relative data columns (fall back to grey for any unseen sequences)
rel_missing_sequences <- setdiff(rel_sequence_names, names(color_dict))
if (length(rel_missing_sequences) > 0) {
  color_dict <- c(color_dict, setNames(rep("#808080", length(rel_missing_sequences)), rel_missing_sequences))
}
rel_color_palette <- color_dict[rel_sequence_names]

seq_rel_long <- seq_relative_abundance %>%
  pivot_longer(cols = -sample_name,
               names_to = "sequence",
               values_to = "abundance")
seq_rel_long$sequence <- factor(seq_rel_long$sequence, levels = seq_order)
seq_rel_long$sample_name <- as.character(seq_rel_long$sample_name)
seq_rel_long$sample_name <- factor(seq_rel_long$sample_name, levels = sample_levels)

# Add site grouping for faceting
seq_rel_long <- seq_rel_long %>%
  mutate(site = ifelse(grepl("^ARM01", sample_name), "Seaward Reef Colonies", "Lagoon Colonies"),
         site = factor(site, levels = c("Seaward Reef Colonies", "Lagoon Colonies")))

p2 <- ggplot(seq_rel_long, aes(x = sample_name, y = abundance, fill = sequence)) +
  geom_bar(stat = "identity") +
  facet_grid(~site, scales = "free_x", space = "free_x", switch = "x") +
  scale_x_discrete(labels = colony_name_map) +
  scale_fill_manual(values = rel_color_palette, na.translate = FALSE) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_pubr(base_size = 14) +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.position = "bottom",
        legend.text = element_text(size = 10),
        legend.title = element_text(face = "bold"),
        legend.key.size = unit(0.4, "cm"),
        legend.box.just = "center",
        legend.margin = margin(0, 0, 0, 0),
        legend.spacing.y = unit(2, "pt"),
        plot.margin = margin(10, 10, 10, 10),
        strip.background = element_blank(),
        strip.text = element_text(size = 16, face = "bold"),
        strip.placement = "outside",
        panel.spacing.x = unit(8, "mm")) +
  guides(fill = guide_legend(ncol = 8, reverse = TRUE, title.position = "top")) +
  labs(x = "", y = "Relative Abundance", fill = "ITS2 Sequence")

print(p2)

# Save plot
ggsave(here("figures/seq_relative_abundance.png"), p2, width = 10, height = 10, dpi = 300)
ggsave(here("figures/seq_relative_abundance.svg"), p2, width = 10, height = 10)
```

## ITS2 Type Profile Absolute Abundance by Sample

```{r}
#| fig-width: 10
#| fig-height: 10
#| dpi: 300

# Load ITS2 type profile absolute abundance data
type_absolute_abundance <- read.csv(
  here("symportal_data/20260116T164045_hoongbe1/its2_type_profiles/its2_type_profiles_absolute.csv"),
  check.names = FALSE
)

# Load SymPortal type profile colour dictionary
type_color_dict <- jsonlite::fromJSON(
  here("symportal_data/20260116T164045_hoongbe1/html/prof_color_dict.json")
)

# Load metadata to map type profile names to UIDs
type_meta <- read.delim(
  here("symportal_data/20260116T164045_hoongbe1/its2_type_profiles/665_20260116T182401_DBV_20260117T155228.profiles.meta_only.txt"),
  check.names = FALSE
)

# Create mapping from type profile name to UID
name_to_uid_mapping <- setNames(
  as.character(type_meta$`ITS2 type profile UID`),
  type_meta$`ITS2 type profile`
)

# Ensure every type profile has a colour; fall back to grey if missing
type_names <- setdiff(names(type_absolute_abundance), "sample_name")
# Convert JSON keys to character for matching (in case they're numeric strings)
type_dict_keys_as_char <- as.character(names(type_color_dict))
missing_types <- setdiff(as.character(type_names), type_dict_keys_as_char)
if (length(missing_types) > 0) {
  type_color_dict <- c(type_color_dict, setNames(rep("#808080", length(missing_types)), missing_types))
}

# Order type profiles by total abundance; reverse to place least abundant at the bottom
type_totals_abs <- colSums(type_absolute_abundance[type_names], na.rm = TRUE)
type_order <- rev(names(sort(type_totals_abs, decreasing = TRUE)))
# Build palette using name-to-UID mapping
type_color_palette <- sapply(type_order, function(x) {
  uid <- name_to_uid_mapping[x]
  if (!is.na(uid) && uid %in% names(type_color_dict)) {
    type_color_dict[[uid]]
  } else {
    "#808080"  # fallback grey
  }
})

# Convert to long format for ggplot2
type_long <- type_absolute_abundance %>%
  pivot_longer(cols = -sample_name, 
               names_to = "type_profile", 
               values_to = "abundance")
type_long$type_profile <- factor(type_long$type_profile, levels = type_order)
type_long$sample_name <- as.character(type_long$sample_name)
type_long$sample_name <- factor(type_long$sample_name, levels = sample_levels)

# Add site grouping for faceting
type_long <- type_long %>%
  mutate(site = ifelse(grepl("^ARM01", sample_name), "Seaward Reef Colonies", "Lagoon Colonies"),
         site = factor(site, levels = c("Seaward Reef Colonies", "Lagoon Colonies")))

# Legend order: first appearance from left to right across samples
legend_order <- type_long %>%
  filter(abundance > 0) %>%
  arrange(sample_name) %>%
  pull(type_profile) %>%
  as.character() %>%
  unique()
if (length(legend_order) == 0) legend_order <- type_order

# Create stacked bar chart
p3 <- ggplot(type_long, aes(x = sample_name, y = abundance, fill = type_profile)) +
  geom_bar(stat = "identity") +
  facet_grid(~site, scales = "free_x", space = "free_x", switch = "x") +
  scale_x_discrete(labels = colony_name_map) +
  scale_fill_manual(values = type_color_palette[legend_order], breaks = legend_order, na.translate = FALSE) +
  scale_y_continuous(expand = c(0, 0), labels = scales::label_number(suffix = "k", scale = 1/1000)) +
  theme_pubr(base_size = 14) +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.position = "bottom",
        legend.text = element_text(size = 10),        legend.title = element_text(face = "bold"),        legend.key.size = unit(0.4, "cm"),
        legend.box.just = "center",
        legend.margin = margin(0, 0, 0, 0),
        legend.spacing.y = unit(2, "pt"),
        plot.margin = margin(10, 10, 10, 10),
        strip.background = element_blank(),
        strip.text = element_text(size = 16, face = "bold"),
        strip.placement = "outside",
        panel.spacing.x = unit(8, "mm")) +
  guides(fill = guide_legend(ncol = 8, reverse = FALSE, title.position = "top")) +
  labs(x = "", y = "Absolute Abundance", fill = "ITS2 Type Profile")

print(p3)

# Save plot
ggsave(here("figures/type_profile_absolute_abundance.png"), p3, width = 10, height = 10, dpi = 300)
ggsave(here("figures/type_profile_absolute_abundance.svg"), p3, width = 10, height = 10)
```

## ITS2 Type Profile Relative Abundance by Sample

```{r}
#| fig-width: 10
#| fig-height: 10
#| dpi: 300

# Load ITS2 type profile relative abundance data
type_relative_abundance <- read.csv(
  here("symportal_data/20260116T164045_hoongbe1/its2_type_profiles/its2_type_profiles_relative.csv"),
  check.names = FALSE
)

# Use the same type profile order and colours for consistency
rel_type_names <- setdiff(names(type_relative_abundance), "sample_name")
# Build palette using name-to-UID mapping
rel_type_color_palette <- sapply(type_order, function(x) {
  uid <- name_to_uid_mapping[x]
  if (!is.na(uid) && uid %in% names(type_color_dict)) {
    type_color_dict[[uid]]
  } else {
    "#808080"  # fallback grey
  }
})

type_rel_long <- type_relative_abundance %>%
  pivot_longer(cols = -sample_name,
               names_to = "type_profile",
               values_to = "abundance")
type_rel_long$type_profile <- factor(type_rel_long$type_profile, levels = type_order)
type_rel_long$sample_name <- as.character(type_rel_long$sample_name)
type_rel_long$sample_name <- factor(type_rel_long$sample_name, levels = sample_levels)

# Add site grouping for faceting
type_rel_long <- type_rel_long %>%
  mutate(site = ifelse(grepl("^ARM01", sample_name), "Seaward Reef Colonies", "Lagoon Colonies"),
         site = factor(site, levels = c("Seaward Reef Colonies", "Lagoon Colonies")))

# Add asterisks to mark bleached colonies
bleached_colonies <- c("ARM01-1", "ARM01-4", "ARM01-5", "ARM01-6")
colony_name_map[bleached_colonies] <- paste0(colony_name_map[bleached_colonies], "*")

p4 <- ggplot(type_rel_long, aes(x = sample_name, y = abundance, fill = type_profile)) +
  geom_bar(stat = "identity") +
  facet_grid(~site, scales = "free_x", space = "free_x", switch = "x") +
  scale_x_discrete(labels = colony_name_map) +
  scale_fill_manual(values = rel_type_color_palette[legend_order], breaks = legend_order, na.translate = FALSE) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1.0)) +
  theme_pubr(base_size = 14) +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.position = "bottom",
        legend.text = element_text(size = 10),        legend.title = element_text(face = "bold"),        legend.key.size = unit(0.4, "cm"),
        legend.box.just = "center",
        legend.margin = margin(0, 0, 0, 0),
        legend.spacing.y = unit(2, "pt"),
        plot.margin = margin(10, 10, 10, 10),
        strip.background = element_blank(),
        strip.text = element_text(size = 16, face = "bold"),
        strip.placement = "outside",
        panel.spacing.x = unit(8, "mm")) +
  guides(fill = guide_legend(ncol = 8, reverse = FALSE, title.position = "top")) +
  labs(x = "", y = "Relative Abundance", fill = "ITS2 Type Profile")

print(p4)

# Save plot
ggsave(here("figures/type_profile_relative_abundance.png"), p4, width = 10, height = 10, dpi = 300)
ggsave(here("figures/type_profile_relative_abundance.svg"), p4, width = 10, height = 10)
```

## Combined Figure: Sequence and Type Profile Absolute Abundance

```{r}
#| fig-width: 10
#| fig-height: 16
#| dpi: 300

# Combine p1 (sequence absolute abundance) and p3 (type profile absolute abundance)
spacer_abs <- cowplot::ggdraw()  # blank spacer between panels

combined_abs_figure <- cowplot::plot_grid(p1, spacer_abs, p3,
                                          nrow = 3, ncol = 1,
                                          rel_heights = c(1, 0.075, 0.8),
                                          labels = c("A", "", "B"),
                                          label_size = 16,
                                          align = "v")

print(combined_abs_figure)

# Save combined figure
ggsave(here("figures/combined_absolute_abundance.png"), combined_abs_figure, width = 10, height = 16, dpi = 300)
ggsave(here("figures/combined_absolute_abundance.svg"), combined_abs_figure, width = 10, height = 16)
```

## Combined Figure: Sequence and Type Profile Relative Abundance

```{r}
#| fig-width: 10
#| fig-height: 16
#| dpi: 300

# Combine p2 (sequence relative abundance) and p4 (type profile relative abundance)
# Create a spacer for more space between panels
spacer <- plot.new()

combined_figure <- cowplot::plot_grid(p2, spacer, p4, 
                                       nrow = 3, ncol = 1,
                                       rel_heights = c(1, 0.075, 0.8),
                                       labels = c("A", "", "B"),
                                       label_size = 16,
                                       align = "v")

print(combined_figure)

# Save combined figure
ggsave(here("figures/combined_relative_abundance.png"), combined_figure, width = 10, height = 16, dpi = 300)
ggsave(here("figures/combined_relative_abundance.svg"), combined_figure, width = 10, height = 16)
```

## Filter ITS2 Sequences

```{r}
# Read sequences to keep from DIV list
div_sequences <- readLines(here("data_analysis/DIV_list.txt"), warn = FALSE)
div_sequences <- gsub("^\\ufeff", "", div_sequences)
div_sequences <- unique(trimws(div_sequences))
div_sequences <- div_sequences[div_sequences != ""]

if (length(div_sequences) == 0) {
  stop("DIV_list.txt is empty. Add sequence names to keep before running this chunk.")
}

available_sequences <- setdiff(names(seq_relative_abundance), "sample_name")
matched_div_sequences <- intersect(div_sequences, available_sequences)

if (length(matched_div_sequences) == 0) {
  stop("No DIV list sequences matched columns in seq_relative_abundance. Check spelling/format in DIV_list.txt.")
}

# Create new dataframe where non-DIV sequences are grouped as "minor"
seq_relative_abundance_filtered <- seq_relative_abundance %>%
  pivot_longer(
    cols = -sample_name,
    names_to = "sequence",
    values_to = "abundance"
  ) %>%
  mutate(sequence = if_else(sequence %in% matched_div_sequences, sequence, "minor")) %>%
  group_by(sample_name, sequence) %>%
  summarise(abundance = sum(abundance, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = sequence, values_from = abundance, values_fill = 0)
```

## Filtered ITS2 Sequence Relative Abundance by Sample

```{r}
#| fig-width: 10
#| fig-height: 10
#| dpi: 300

filtered_sequence_names <- setdiff(names(seq_relative_abundance_filtered), "sample_name")

# Build color palette for filtered sequences and add grey for minor group
filtered_missing_sequences <- setdiff(filtered_sequence_names, names(color_dict))
filtered_missing_sequences <- setdiff(filtered_missing_sequences, "minor")
if (length(filtered_missing_sequences) > 0) {
  color_dict <- c(color_dict, setNames(rep("#808080", length(filtered_missing_sequences)), filtered_missing_sequences))
}
filtered_color_palette <- color_dict[setdiff(filtered_sequence_names, "minor")]
filtered_color_palette <- c(filtered_color_palette, minor = "#808080")

filtered_seq_long <- seq_relative_abundance_filtered %>%
  pivot_longer(cols = -sample_name,
               names_to = "sequence",
               values_to = "abundance") %>%
  mutate(sequence = if_else(sequence == "minor", "Filtered sequences", sequence),
         sample_name = as.character(sample_name),
         sample_name = factor(sample_name, levels = sample_levels),
         site = ifelse(grepl("^ARM01", sample_name), "Seaward Reef Colonies", "Lagoon Colonies"),
         site = factor(site, levels = c("Seaward Reef Colonies", "Lagoon Colonies")))

filtered_sequence_order <- filtered_seq_long %>%
  group_by(sequence) %>%
  summarise(total = sum(abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(total) %>%
  pull(sequence) %>%
  as.character()

# Reorder so minor is at bottom (first in factor levels for stacking)
if ("Filtered sequences" %in% filtered_sequence_order) {
  filtered_sequence_order <- c(
    "Filtered sequences",
    setdiff(filtered_sequence_order, "Filtered sequences")
  )
}

filtered_seq_long$sequence <- factor(filtered_seq_long$sequence, levels = filtered_sequence_order)

filtered_fill_values <- setNames(
  unname(filtered_color_palette[ifelse(filtered_sequence_order == "Filtered sequences", "minor", filtered_sequence_order)]),
  filtered_sequence_order
)

filtered_legend_order <- filtered_seq_long %>%
  filter(abundance > 0) %>%
  arrange(sample_name) %>%
  pull(sequence) %>%
  as.character() %>%
  unique()
if (length(filtered_legend_order) == 0) filtered_legend_order <- filtered_sequence_order
if ("Filtered sequences" %in% filtered_legend_order) {
  filtered_legend_order <- c(
    setdiff(filtered_legend_order, "Filtered sequences"),
    "Filtered sequences"
  )
}

p5 <- ggplot(filtered_seq_long, aes(x = sample_name, y = abundance, fill = sequence)) +
  geom_bar(stat = "identity") +
  facet_grid(~site, scales = "free_x", space = "free_x", switch = "x") +
  scale_x_discrete(labels = colony_name_map) +
  scale_fill_manual(values = filtered_fill_values[filtered_legend_order], breaks = filtered_legend_order, na.translate = FALSE) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_cartesian(ylim = c(0, 1.0)) +
  theme_pubr(base_size = 14) +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.position = "bottom",
        legend.text = element_text(size = 10),
        legend.title = element_text(face = "bold"),
        legend.key.size = unit(0.4, "cm"),
        legend.box.just = "center",
        legend.margin = margin(0, 0, 0, 0),
        legend.spacing.y = unit(2, "pt"),
        plot.margin = margin(10, 10, 10, 10),
        strip.background = element_blank(),
        strip.text = element_text(size = 16, face = "bold"),
        strip.placement = "outside",
        panel.spacing.x = unit(8, "mm")) +
  guides(fill = guide_legend(ncol = 12, reverse = FALSE, title.position = "top")) +
  labs(x = "", y = "Relative Abundance", fill = "ITS2 Sequence")

print(p5)

# Save plot
ggsave(here("figures/seq_relative_abundance_filtered.png"), p5, width = 10, height = 10, dpi = 300)
ggsave(here("figures/seq_relative_abundance_filtered.svg"), p5, width = 10, height = 10)
```

## Combined Figure: Filtered Sequences and Type Profile Relative Abundance

```{r}
#| fig-width: 10
#| fig-height: 16
#| dpi: 300

# Combine p5 (filtered sequence relative abundance) and p4 (type profile relative abundance)
# Create a spacer for more space between panels
spacer <- plot.new()

combined_filtered_figure <- cowplot::plot_grid(p5, spacer, p4, 
                                               nrow = 3, ncol = 1,
                                               rel_heights = c(1, 0.075, 1),
                                               labels = c("A", "", "B"),
                                               label_size = 16,
                                               align = "v")

print(combined_filtered_figure)

# Save combined figure
ggsave(here("figures/combined_filtered_and_type_relative_abundance.png"), combined_filtered_figure, width = 10, height = 16, dpi = 300)
ggsave(here("figures/combined_filtered_and_type_relative_abundance.svg"), combined_filtered_figure, width = 10, height = 16)
```

## Combined Figure: Filtered Sequences and Type Profile Relative Abundance (Side-by-side)

```{r}
#| fig-width: 16
#| fig-height: 7
#| dpi: 300

# Combine p5 and p4 side-by-side
combined_filtered_figure_sidebyside <- cowplot::plot_grid(p5, spacer, p4,
                                                           nrow = 1, ncol = 3,
                                                           rel_widths = c(1, 0.05, 1),
                                                           labels = c("A", "", "B"),
                                                           label_size = 16,
                                                           align = "v")

print(combined_filtered_figure_sidebyside)

# Save combined figure
ggsave(here("figures/combined_filtered_and_type_relative_abundance_sidebyside.png"), combined_filtered_figure_sidebyside, width = 16, height = 7, dpi = 300)
ggsave(here("figures/combined_filtered_and_type_relative_abundance_sidebyside.svg"), combined_filtered_figure_sidebyside, width = 16, height = 7)
```

# DIV Data Summary

```{r}
# Count all detected DIVs in each sample from seq_absolute_abundance
all_div_columns <- setdiff(names(seq_absolute_abundance), "sample_name")

# Optional comparison: count only DIVs listed in DIV_list.txt
div_sequences_local <- readLines(here("data_analysis/DIV_list.txt"), warn = FALSE)
div_sequences_local <- gsub("^\\ufeff", "", div_sequences_local)
div_sequences_local <- unique(trimws(div_sequences_local))
div_sequences_local <- div_sequences_local[div_sequences_local != ""]

div_list_columns <- intersect(div_sequences_local, all_div_columns)

div_presence_summary <- seq_absolute_abundance %>%
  transmute(
    sample_name,
    n_DIVs_present_all = rowSums(across(all_of(all_div_columns), ~ .x > 0), na.rm = TRUE),
    n_DIVs_present_from_DIV_list = if (length(div_list_columns) > 0) {
      rowSums(across(all_of(div_list_columns), ~ .x > 0), na.rm = TRUE)
    } else {
      NA_integer_
    }
  ) %>%
  arrange(sample_name)

avg_divs_per_sample_all <- mean(div_presence_summary$n_DIVs_present_all, na.rm = TRUE)

if (length(div_list_columns) > 0) {
  avg_divs_per_sample_div_list <- mean(div_presence_summary$n_DIVs_present_from_DIV_list, na.rm = TRUE)
}

print(div_presence_summary)
cat(sprintf("Average number of DIVs per sample (all detected): %.2f\n", avg_divs_per_sample_all))

if (length(div_list_columns) > 0) {
  cat(sprintf("Average number of DIVs per sample (DIV_list only): %.2f\n", avg_divs_per_sample_div_list))
} else {
  cat("No DIV_list.txt entries matched seq_absolute_abundance columns. DIV_list-only summary skipped.\n")
}
```

# Bleaching Status Venn Diagram
```{r}
bleaching_data <- read.csv(here("symportal_data/bleaching_data.csv"))

bleaching_sets <- bleaching_data %>%
  filter(!is.na(bleaching_status), !is.na(dominant_symbiont)) %>%
  mutate(
    bleaching_status = trimws(as.character(bleaching_status)),
    dominant_symbiont = trimws(as.character(dominant_symbiont))
  ) %>%
  distinct(bleaching_status, dominant_symbiont) %>%
  group_by(bleaching_status) %>%
  summarise(dominant_symbiont = list(sort(unique(dominant_symbiont))), .groups = "drop") %>%
  tibble::deframe()

if (!all(c("bleached", "unbleached") %in% names(bleaching_sets))) {
  stop("Expected both 'bleached' and 'unbleached' bleaching_status values in bleaching_data.csv.")
}

bleached_symbionts <- bleaching_sets[["bleached"]]
unbleached_symbionts <- bleaching_sets[["unbleached"]]

bleached_only <- setdiff(bleached_symbionts, unbleached_symbionts)
unbleached_only <- setdiff(unbleached_symbionts, bleached_symbionts)
shared_symbionts <- intersect(bleached_symbionts, unbleached_symbionts)

circle_points <- function(center_x, center_y, radius, n = 200) {
  angle <- seq(0, 2 * pi, length.out = n)
  data.frame(
    x = center_x + radius * cos(angle),
    y = center_y + radius * sin(angle)
  )
}

bleached_circle <- circle_points(0, 0.65, 1)
unbleached_circle <- circle_points(0, -0.65, 1)

bleaching_venn <- ggplot() +
  geom_polygon(
    data = bleached_circle,
    aes(x = x, y = y),
    fill = "#ffbf00",
    color = "#ffbf00",
    alpha = 0.55,
    linewidth = 1.2
  ) +
  geom_polygon(
    data = unbleached_circle,
    aes(x = x, y = y),
    fill = "#92d051",
    color = "#92d051",
    alpha = 0.55,
    linewidth = 1.2
  ) +
  annotate("text", x = 0, y = 1, label = length(bleached_only), size = 6, fontface = "bold") +
  annotate("text", x = 0, y = 0, label = length(shared_symbionts), size = 6, fontface = "bold") +
  annotate("text", x = 0, y = -1, label = length(unbleached_only), size = 6, fontface = "bold") +
  annotate("text", x = 0, y = 1.75, label = "Profiles in Bleached Colonies", color = "black", fontface = "bold", size = 5, vjust = 0) +
  annotate("text", x = 0, y = -1.75, label = "Profiles in Unbleached Colonies", color = "black", fontface = "bold", size = 5, vjust = 1) +
  coord_equal(xlim = c(-1.55, 2.2), ylim = c(-1.9, 1.9), clip = "off") +
  theme_void(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    plot.margin = margin(20, 20, 20, 20)
  )

print(bleaching_venn)

ggsave(here("figures/bleaching_status_venn_diagram.png"), bleaching_venn, width = 8, height = 6, dpi = 300)
ggsave(here("figures/bleaching_status_venn_diagram.svg"), bleaching_venn, width = 8, height = 6)


```